---
title: "Getting Started with `semidentify`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{semidentify-quickstart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(semidentify)
```

## Introduction

`semidentify` provides rule-based diagnostic checks for selected
identification conditions in structural equation models (SEMs). The package
can evaluate models supplied as `lavaan` syntax, `lavaan` parameter tables, or
fitted `lavaan` objects.

The package is intended as a quick diagnostic tool. Passing all applicable
necessary rules does not, by itself, establish identification. Conversely, a
satisfied sufficient rule should be interpreted according to the assumptions
of that rule.

## Basic workflow

The primary function is `id()`. Supply a model in `lavaan` syntax and specify
the `lavaan` function whose defaults you intend to use through `lav_fun`.

```{r basic-id}
model <- '
  L1 =~ Y1 + Y2 + Y3
  L2 =~ Y4 + Y5 + Y6
  L2 ~ L1
'

id(model, lav_fun = "sem")
```

The output table reports whether each implemented rule passes and whether the
rule is necessary, sufficient, or neither for the relevant model class.
When `print_msgs = TRUE`, explanatory messages are printed below the table.

```{r no-messages}
id(model, lav_fun = "sem", print_msgs = FALSE)
```

### Interpreting messages

When `print_msgs = TRUE`, `semidentify` prints explanatory messages below the
rule-check table. The numbers in the table's `Message` column correspond to
numbered messages in this section.

Messages are assigned one of three levels:

- **Info**: The rule is not applicable to the current model, or the package is
  providing contextual information about how the rule was evaluated.
- **Reason**: A condition of a sufficient rule was not met. This does not by
  itself establish that the model is underidentified; it means that the rule
  alone cannot establish identification for the model.
- **WARNING**: A necessary condition was not met. Under the assumptions of the
  relevant rule, this indicates that the model is *not* identified.

For example, a `Reason` message for a sufficient rule should be interpreted
differently from a `WARNING` message for a necessary rule. A `Reason` means
that the package **could not use that particular sufficient condition** to certify
identification. A `WARNING` identifies a potential identification problem that
**requires model respecification** or further investigation.


## Choosing `lav_fun`

When a model string is supplied, `lav_fun` determines which `lavaan` defaults
are used when the model is converted to a parameter table. Current options
are:

- `"lavaan"`
- `"sem"`
- `"cfa"`

For example, use `"cfa"` for a confirmatory factor analysis model:

```{r cfa}
cfa_model <- '
  L1 =~ Y1 + Y2 + Y3
  L2 =~ Y4 + Y5 + Y6
  L1 ~~ L2
'

id(cfa_model, lav_fun = "cfa")
```

When a parameter table or fitted `lavaan` object is supplied, its constraints
have already been established, so `lav_fun` does not affect the rule checks.

## Checking latent-variable scaling

Use `scaling()` to inspect whether each latent variable has scaling
constraints commonly used in the literature.

```{r scaling}
scaling(model, lav_fun = "sem")
```

For models without a mean structure, the package checks whether each latent
variable has assigned units through a fixed nonzero loading or a fixed positive
latent-variable variance.

For models with a mean structure, the package also checks whether the latent
variable has an assigned origin, such as through a fixed latent-variable mean
or a fixed intercept for a scaling indicator.

To return only a logical result for each latent variable:

```{r scaling-logical}
scaling(
  model,
  lav_fun = "sem",
  return_type = "logical"
)
```

## Working with a fitted `lavaan` model

If a model has already been fitted with `lavaan`, pass the fitted object
directly to `id()` or `scaling()`. Set `lav_fun = NA` to avoid warnings
when `lav_fun` does not match the original call.

```{r fitted-model, eval = FALSE}
library(lavaan)

fit <- sem(model, data = my_data)

id(fit, lav_fun = NA)
scaling(fit, lav_fun = NA)
```

## Piping identification and scaling checks

`semidentify` supports a sequential diagnostic workflow. For example, the
following first prints the identification-rule results and then prints the
scaling results:

```{r piping}
id(model, lav_fun = "sem") |> scaling()
```

The reverse order is also supported:

```{r reverse-piping}
scaling(model, lav_fun = "sem") |> id(print_msgs = FALSE)
```

## The two-step rule

For supported full SEMs, `id2()` applies the package's two-step procedure. It
creates a measurement-model transformation and a latent-variable structural
transformation, then evaluates identification rules for both components.

```{r two-step}
id2(model, lav_fun = "sem")
```

The two-step rule is currently intended for supported single-level SEMs and is
not applicable to every model type.

## Next steps

For details on individual rules, see:

```{r help, eval = FALSE}
?id
?scaling
?get_rules
```

You can retrieve implemented rule functions with `get_rules()`:

```{r get-rules}
cfa_rules <- get_rules(rule = "*", model_type = "cfa")
```

For further theoretical background, see Ken Bollen's book *Elements of Structural Equation Models* (2026).
