---
title: "Getting Started with `semrulesid`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{semrulesid-quickstart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(semrulesid)
```

## Introduction

`semrulesid` allows the user to check a structural equation model (SEM) written
in [`lavaan`](https://lavaan.ugent.be/) (Rosseel, 2012) syntax against a number
of identification rules from the literature. Rules are specified as being
necessary and/or sufficient and specific reasons are given when a rule is broken
or not applicable. Users should not treat the package output as the sole
determinant of model identification. Instead, `semrulesid` should be used as a
quick check for potential identification issues and outstanding
model-specification concerns.


## Basic workflow

The primary function is `id()`. Supply a model in `lavaan` syntax and specify
the `lavaan` function whose defaults you intend to use through `lav_fun`.

```{r basic-id}
model <- '
  L1 =~ Y1 + Y2 + Y3
  L2 =~ Y4 + Y5 + Y6
  L2 ~ L1
'

id(model, lav_fun = "sem")
```

The output table reports whether each implemented rule passes and whether the
rule is necessary, sufficient, or neither for the relevant model class, which is
printed at the top of the output.

### Interpreting messages

The `id()` function prints explanatory messages below the rule-check table. The
numbers in the table's `Message` column correspond to numbered messages in this
section.

Messages are assigned one of three levels:

- **Info**: The rule is not applicable to the current model or model type, or
  the package is providing contextual information about how the rule was
  evaluated.
- **Reason**: A condition of a *sufficient* rule was not met. This does not by
  itself establish that the model is underidentified; it means that the rule
  alone cannot establish identification for the model.
- **WARNING**: A *necessary* condition was not met. Under the assumptions of the
  relevant rule, this indicates that the model is *not* identified.

A `Reason` message for a sufficient rule should be interpreted differently from
a `WARNING` message for a necessary rule. A `Reason` means that the package
*could not use that particular sufficient condition* to certify identification.
A `WARNING` identifies a potential identification problem that *requires model
respecification* or further investigation.

## Choosing `lav_fun`

When a model string is supplied, `lav_fun` determines which `lavaan` defaults
are used when the model is converted to a parameter table. Current options are:

- `"lavaan"`
- `"sem"`
- `"cfa"`

For example, use `"cfa"` for a confirmatory factor analysis model:

```{r cfa, eval=FALSE}
cfa_model <- '
  L1 =~ Y1 + Y2 + Y3
  L2 =~ Y4 + Y5 + Y6
  L1 ~~ L2
'

id(cfa_model, lav_fun = "cfa")
```

When a parameter table or fitted `lavaan` object is supplied, `lav_fun` is
ignored since the model defaults have already been applied.

## Checking latent-variable scaling

Use `scaling()` to inspect whether each latent variable scaled via a method used
in the literature.

```{r scaling}
scaling(model, lav_fun = "sem")
```

For models without a mean structure, the package checks whether each latent
variable has assigned units through a fixed, nonzero loading (scaling indicator)
or a fixed positive latent-variable variance.

For models with a mean structure, the package also checks whether the latent
variable has an assigned origin, such as through a fixed latent-variable mean or
a fixed intercept for a scaling indicator.

## Working with a fitted `lavaan` model

If a model has already been fitted with `lavaan`, pass the fitted object
directly to `id()` or `scaling()`. Set `lav_fun = NA` to avoid warnings when
`lav_fun` does not match the call used to fit the model.

```{r fitted-model, eval = FALSE}
library(lavaan)

fit <- sem(model, data = my_data)

id(fit, lav_fun = NA)
scaling(fit, lav_fun = NA)
```

## Piping identification and scaling checks

`semrulesid` supports the use of a pipe operator to chain together
identification and scaling checks, e.g., using the base R pipe `|>`:

```{r piping, eval=FALSE}
id(model, lav_fun = "sem") |> scaling()
```

The reverse order is also supported:

```{r reverse-piping,eval=FALSE}
scaling(model, lav_fun = "sem") |> id(print_msgs = FALSE)
```

## The two-step rule

For supported full SEMs (latent variables plus structural paths), `id2()`
applies the two-step rule of identification (Bollen, 2026). It first transforms
the model into a confirmatory factor analysis (CFA) model and evaluates it for
identification. It then transforms the model into a simultaneous equations
model, and evaluates *it* for identification. If both steps are identified, the
original model is identified. The outputs of both steps are printed via the
`id2()` function.

```{r two-step}
id2(model, lav_fun = "sem")
```

The two-step rule is currently intended for supported single-level SEMs and is
not applicable to every model type.

## Next steps

For details on individual rules, see:

```{r help, eval = FALSE}
?id
?scaling
?get_rules
```

You can retrieve implemented rule functions with `get_rules()`:

```{r get-rules}
cfa_rules <- get_rules(rule = "*", model_type = "cfa")
```

For further theoretical background, see Ken Bollen's book *Elements of
Structural Equation Models* (2026).
